---
title: "create_annotation"
author: "Victor Yuan"
date: "02/08/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

Outline

1. Get illumina provided cpg and genome locations
2. Use annotatr to provide 'basic' annotations, such as cpg island, gene elements, enhacners
3. Use published datasets to provide other interesting annotation info (e.g. imprinting)
4. Use rtracklayer::liftOver to map from build hg19 to build hg38 genome locations

# Libraries 

```{r}
library(rtracklayer) 
library(annotatr)
library(tidyverse)
library(readxl)
library(fuzzyjoin)
library(janitor)
```

# 1. Illumina annotation

We start with the illumina provided locations of each probe. Note that Illumina has released multiple versions of the EPIC, each with a different set of probes (200-2000 probes can be different between arrays).

 Here I combine the b4 annotation with those probes that were removed from B2 to B3 versions. This set of probes
will be a starting point for cpgs to annotate.

TODO: Add in b2 and b1 annotations to ensure all EPIC probes in the universe will be covered.
TODO: Add in 450k probes
TODO: Add in 27k probes?

```{r}
b4 <- read_csv('Z:/Victor/Data/DNAm annotations/MethylationEPIC_v-1-0_B4.csv', skip = 7)
b4_missing <- read_csv('Z:/Victor/Data/DNAm annotations/MethylationEPIC Missing Legacy CpG (v1.0_B3 vs. v1.0_B2) Annotations.csv')
anno <- bind_rows(b4, b4_missing %>% mutate_at(c('AddressA_ID', 'AddressB_ID'), as.character))
```

# 2. Use annotatr to add UCSC annotations, and FANTOM5 enhancers

The gather and spread functions take 10 hours to complete.

First I filter to the illumina information to keep. 

Except for Start, End, chr and cpg_id columns, I rename others with the prefix 'ilmn', to indicate 
they are from the illumina annotation file. People can refer to the illumina files for description 
of those variables

```{r}
# clean up annotation
anno <- anno %>% mutate(Strand = ifelse(Strand == 'F', '+', '-'),
                start = MAPINFO,
                end = MAPINFO,
                chr = paste0('chr', CHR),
                cpg_id = Name) %>%
  
  # keep basic information
  dplyr::select(cpg_id, chr, start, end, Strand, AddressA_ID:Forward_Sequence) %>%
  
  # remove rs, and control probes
  filter(grepl('^[(cg)(ch)]{1}', cpg_id)) %>%
  
  # add illumina identifier
  rename_at(vars(-(1:5)), ~paste0('ilmn_', .))

# coerce to granges
anno_GR <- makeGRangesFromDataFrame(anno, keep.extra.columns = T)

# select annotations to use
annots <- c('hg19_enhancers_fantom',
            'hg19_cpgs', 
            'hg19_genes_1to5kb',
            'hg19_genes_promoters', 
            'hg19_genes_5UTRs',
            'hg19_genes_exons',
            'hg19_genes_intronexonboundaries',
            'hg19_genes_introns',
            'hg19_genes_3UTRs',
            'hg19_genes_intergenic')
annots <- build_annotations(genome = 'hg19', annotations = annots)

# annotate the base annotation
annotated <- annotate_regions(regions = anno_GR, annotations = annots, ignore.strand = T) %>% 
  as_tibble

# clean it up
annotated
x1 <- length(unique(annotated$cpg_id))

system.time(
x <- annotated %>% 
  
  # 1. remove identifiers from annot.id, i.e "intron.9999" should be just called "intron"
  mutate(annot.id = str_extract(annot.id, '^[A-z0-9]+'),
         annot.type = str_extract(annot.type, '(?<=hg19_)[^_]+')
         ) %>%
  dplyr::select(-annot.seqnames, -annot.start, -annot.end, -annot.strand) %>%
  
  # 2. paste together all elements, grouped by cpg-related and gene-related elements
  ### I'm not sure but this step might be able to skip
  group_by(cpg_id,annot.type) %>%
  summarize(
    # paste(unique(.)) ensures all nonunique mappings are retained
    cpg = paste(unique(cpg_id)),
    chr = paste(unique(seqnames)),
    start = paste(unique(start)),
    end = paste(unique(end)),
    
    id = paste(annot.id, collapse = ", "),
    width = paste(annot.width, collapse = ", "),
    tx_id = paste(annot.tx_id, collapse = ", "),
    gene_id = paste(annot.gene_id, collapse = ", "),
    symbol = paste(annot.symbol, collapse = ", "))  %>%
  
  # 3. now spread multiple values across columns
  # first create a temporary variable to spread by, using gather() and unite()
  ungroup() %>%
  dplyr::select(-cpg_id)  %>%
  gather(variable, value, -(annot.type:end)) %>%
  unite(temp, annot.type, variable, sep = '_') %>%
  spread(temp, value) %>%
  
  # 4. clean up
  # remove uninformative columns
  dplyr::select(-enhancers_gene_id , -enhancers_symbol, -enhancers_tx_id, 
                -cpg_gene_id, -cpg_symbol, -cpg_tx_id) %>%
  
  # cpg_id == inter replace with "sea"
  mutate(cpg_id = ifelse(cpg_id == 'inter', 'sea', cpg_id)) %>%
  mutate_all(~replace(., . == 'NA', NA_character_))
) # takes 10 hours
# ensure the number of cpgs ended up with equals the number of unique cpgs starting out with
x1 == nrow(x) # must be T 

# now add back in the illumina columns
  
```
saveRDS(x, '../../data/main/interim/annotation.rds')
write_csv(x, '../../data/main/processed/annotation.csv')

x <- readRDS('Z:/Victor/Projects/NIH - cells/data/main/interim/annotation.rds')

# 3. Add other annotations

PMDs

https://www.pnas.org/content/110/15/6037

Schroeder DI, Blair JD, Lott P, Yu HOK, Hong D, Crary F, et al. The human placenta methylome. PNAS. 2013 Apr 9;110(15):6037–42. 

```{r}
# load pmds
pmds <- read_xls('Z:/Victor/Projects/NIH - cells/data/external/placental_pmds.xls', col_names = F)
colnames(pmds) <- c('chr', 'start', 'end')

# join pmds using fuzzyjoin::genome_left_join
anno <- x %>%
  mutate_at(c('start', 'end'), as.numeric) %>%
  genome_left_join(pmds) %>%
  
  # pmd ID is the genome location
  mutate(pmd_id = ifelse(!is.na(chr.y), paste0(chr.y, ':', start.y, '-', end.y), NA),
         pmd_width = end.y - start.y) %>%
  dplyr::rename(chr = chr.x, start = start.x, end = end.x) %>%
  select(-contains('.y'))
```

Imprinting

Created by Giulia

```{r}
imprint <- read_xlsx('Imprinted gene database.xlsx', col_names = T, shee = 'Imprinted genes and DMRs')

# non-placental specific imprinted genes
imprint %>% clean_names %>%
  
  
  
  
  
  
  select(gene, contains('location')) %>% 
  
  
  
  
  
  separate(dmr_location, into = c('dmr_location_1', 'dmr_location_2', 'dmr_location_3', 
                                  'dmr_location_4', 'dmr_location_5'), 
           sep = '[\\s]*[;][\\s]*', fill = 'right') %>%
  gather(key = 'dmr_location_k', value = 'dmr_location_v', dmr_location_1:dmr_location_5,
         na.rm = F) %>%
  arrange(chr, start)  %>% 
  distinct(gene_location, dmr_location_v, .keep_all = T) %>% View()
  
  
    mutate(in_gene = ifelse(grepl('Gene', location_type), 'Gene', NA),
         in_dmr = ifelse(grepl('DMR', location_type), 'DMR', NA)) %>%
  separate(gene_location, into = c('gene_chr', 'gene_start', 'gene_end'), convert = T, remove = F) 
  
  
  
  separate(dmr_location_v, into = c('dmr_chr', 'dmr_start', 'dmr_end'), fill = 'right',
           remove = F) %>%
  dplyr::rename(imprinted_gene_location = gene_location, imprinted_dmr_location = dmr_location_v) %>%
  select(-dmr_location_k, -chr, -end, -start) %>%
  arrange(dmr_chr, dmr_start) %>% View()
  

# non placenta-specific imprints
anno

# placenta-specific


```

# 4. Map from hg19 to hg38

I skip this step for now, since it results in a loss of some cpgs.

```{r eval = F}
# load chain file
chain <- import.chain("hg19ToHg38.over.chain")
chain

# clean anno
anno <- anno %>%
  # rename so coercible to GRANGES 
  dplyr::rename(cpg_id = Name, chr = CHR, start = MAPINFO) %>% 
  mutate(end = start, Strand = ifelse(Strand == 'F', '+', '-')) %>%
  
  # drop columns not interested in
  dplyr::select(cpg_id, chr, start, end, Strand, AddressA_ID:Forward_Sequence) %>%
  
  # remove rs, and control probes
  filter(grepl('^[(cg)(ch)]{1}', cpg_id)) %>%
  makeGRangesFromDataFrame(keep.extra.columns = T)

seqlevelsStyle(anno) <- 'UCSC'

anno38 <- liftOver(anno, chain) %>% unlist

length(anno) - length(anno38) # loses 247 loci
setdiff(anno$cpg_id, anno38$cpg_id) # these specific loci
```


# Column descriptions

```{r}
## column descriptors
colnames(x)
# cpg: cpg identifier
# chr: chromosome position
# start, end: cpg position in bp 
# cpg_id: relation to cpg island features. uses UCSC cpg islands as reference, then definitions for shelves, shores, and seas are used
# cpg_width: width in bp of the cpg island -related feature
# enhancers_id: fantom5 enhancer
# enhancers_width: width in bp of the fantom5 enhancer
# genes_: columns are retrieved from txdb.hsapiens.ucsc.hg19.knowngene
# genes_gene_id: I'm not sure what this is TBH
# genes_id: gene-related elemtn, e.g. promoters, introns/exons, based off of UCSC
# genes_symbol: Gene symbol
# genes_tx_id: transcript identifier
# genes_width: width in bp of gene element
```